A note on observation processes in epidemic models

### 2019-07-30 ms/ms.tex
more tweaks

We assumed that newly infected cases are reported instantaneously when
individuals become infected or when they recover. Although neither of these
assumptions is realistic, they are broadly consistent with the assumptions
of other state-of-art epidemic models (see Introduction). 
More realistic models may distinguish reported and unreported
(i.e., identified and unidentified) cases by adding
new state variables \citep{browne2015modeling,webb2015model} 
or by modeling an explicit delay kernel in reporting time 
\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
birrell2011bayesian, funk2018real}.

While we only considered observation processes associated with incidence 
reporting in this study, it is equally as important to consider observation
processes in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
This assumption does not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: delayed reporting is expected to decrease
the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1). 
Ignoring delays in reporting may affect conclusions about the underlying 
transmission mechanisms.

>>>>

We compared two scenarios in which newly infected cases are reported 
instantaneously (1) when individuals become infected or
(2) when they recover. Although neither of these
assumptions is realistic, many epidemic models still rely on these
assumptions (see Introduction). 
More realistic models may distinguish reported and unreported
(i.e., identified and unidentified) cases by adding
new state variables \citep{browne2015modeling,webb2015model} 
or by modeling an explicit delay kernel in reporting time 
\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
birrell2011bayesian, funk2018real}.

We considered observation processes associated with incidence 
reporting; however, we expect observation processes to be just as important
in mortality reporting. In particular, many disease modelers have
tried to infer underlying transmission mechanisms from historical mortality 
data but assumed that individual deaths are recorded as soon as individuals die 
\citep{he2013inferring, didelot2017model, dean2018human}; this includes
\cite{kermack1927contribution} who approximated the reported \emph{number} of
deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
These frameworks do not account for the possibility 
that a delay in reporting of deaths can
change the shape of an epidemic curve: delays in case reports can decrease
the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1).

There is some empirical evidence that delays in case notification alters the
shape of an epidemic curve and affects our understanding of disease transmission:
(1) Postal delays in measles notification during holidays lead to sudden
epidemic peaks after holidays \citep{fine1982measles}; 
(2) Delays in reports of the foot-and-mouth infection affects the effectiveness
of slaughter policies on epidemic intervention;
and (3) Date of wills written and date of wills probated during
the plague epidemics in London give completely different looking epidemic curves
due to extremely long delays (mean of 2.9 years) between the two \citep{bushby2019wills}.
Ignoring such delays in reporting may affect conclusions about the underlying 
transmission mechanisms.
